---
title: "Building a classifier to predicting house age"
author: "Syed Faizan Aamir"
format: 
  html:
    embed-resources: true

---

## Introduction

This report investigates patterns in residential energy performance across the United Kingdom using supervised machine learning techniques. The study develops and evaluates classification models to predict whether a property was constructed before or after 1930 based on its energy performance characteristics. Three classification methods are considered: Logistic Regression, Linear Discriminant Analysis (LDA), and Quadratic Discriminant Analysis (QDA).

The analysis uses variables derived from Energy Performance Certificate (EPC) data, including current energy efficiency ratings, environmental impact scores, energy consumption, carbon dioxide emissions, and household operating costs. These variables provide measurable indicators of building performance that may reflect differences between older and newer housing stock.

The objective of this report is to compare the predictive performance of the three supervised learning models and determine how effectively energy-related variables distinguish properties built before and after 1930. Model performance is evaluated using classification metrics, confusion matrices, and Receiver Operating Characteristic (ROC) analysis to identify the approach that provides the most reliable predictions.

## The Problem

Data on [Energy Certificates](https://epc.opendatacommunities.org/) were downloaded in March 2025. These certificates are a non-random sample of UK housing, and contain information such as:

- Property energy efficiency: The core data is the energy efficiency rating, which ranges from A (most efficient) to G (least efficient).
- CO2 emissions: The data includes information on the building's likely carbon dioxide emissions.
- Energy costs: Potential costs for heating and lighting the property are provided.
- Improvement recommendations: Each EPC lists cost-effective ways to improve the building's energy performance.

In all, there are 93 variables available on a wide range of metrics. The full dataset is in the course database server at `10.121.4.250` in `energy.energy_certificates`

The data have been used for various published supervised classification exercises:

- ONS data science campus EPC data to predict energy efficiency ratings across Welsh properties. They applied models like XGBoost and found that features such as floor area, number of habitable rooms, and extension count were most predictive <https://datasciencecampus.ons.gov.uk/projects/using-machine-learning-to-predict-energy-efficiency/>
- EPC anomaly detection (improve data quality) <https://www.iaarc.org/publications/fulltext/102_A_Comprehensive_Approach_for_Automated_Anomaly_Detection_and_Enhancement_of_EPC_Datasets_for_Decarbonization.pdf>
- Identifying priorities for retrofit strategy <https://uwlpress.uwl.ac.uk/newvistas/article/id/299/>
- EPC prediction <https://github.com/yourunclefrankie/Energy-efficiency-prediction>

In reducing the carbon footprint of UK housing, it is important to develop a strategy that offers the right solutions to the right properties. It is widely believed that houses build before 1930 require very different approaches to houses built after that period.  In this task, you are required to determine whether it is possible to determine the age of a house based on several recorded energy metrics. These are `current_energy_efficiency`, `environment_impact_current`, `energy_consumption_current`, `co2_emissions_current`, `lighting_cost_current`, `heating_cost_current`, `hot_water_cost_current`.  You have data for all certificates issued in a single local authority area.  This would help the local authority identify the right targets just using metrics rather than requiring a visit to the property to determine its age.


```{r}
#| include: false
reticulate::use_python(
  "/opt/anaconda3/envs/python-exercises/bin/python",
  required = TRUE
)
```

## The Data

A very simple preview of the data is given below:

```{python}
#| echo: false
import pandas as pd
```

And a clumsy presentation of some group summary statistics:


```{python}
#| echo: false
#| results: 'asis'
df = pd.read_csv('../../data_cache/vignettes/supervised_classification/frequencies.csv')
print(df.round(3).to_markdown(index=False))
```



```{python}
#| echo: false
#| results: 'asis'
df = pd.read_csv('../../data_cache/vignettes/supervised_classification/grouped_stats.csv')
print(df.round(3).to_markdown(index=False))
```


The dataset contains 67,087 properties, consisting of 36,523 Pre-1930s properties and 30,464 Post-1930s properties. The class distribution is therefore reasonably balanced, although the older-property group is slightly larger. This balance is useful for supervised classification because it reduces the risk that model performance will be driven mainly by one dominant category.

The descriptive statistics show clear differences between the two construction periods. Post-1930s properties have a higher mean current energy-efficiency score of 63.564, compared with 51.656 for Pre-1930s properties. Their median efficiency score is also higher, at 66 compared with 54, indicating that the difference is not caused only by a small number of extreme values. The spread is slightly lower for Post-1930s homes, with a standard deviation of 11.028, compared with 13.456 for Pre-1930s homes. This suggests that newer properties are not only more efficient on average, but also somewhat more consistent in their energy performance.

A similar pattern appears in the environmental-impact scores. Post-1930s properties have a mean score of 60.501, whereas Pre-1930s properties have a lower mean of 47.304. The corresponding medians are 62 and 48. Since higher environmental-impact scores generally represent stronger environmental performance, these results suggest that newer homes tend to produce better sustainability outcomes.

Energy consumption also differs considerably between the two groups. Pre-1930s properties have a mean current energy-consumption value of 374.525, compared with 272.404 for Post-1930s properties. The median values, 353 and 254, show the same pattern. Carbon dioxide emissions are also higher among older homes, with a mean of 6.782, compared with 4.342 for newer homes. These differences are consistent with the lower energy-efficiency and environmental-impact scores observed for the Pre-1930s group.

Operating costs show mixed patterns. Heating costs display a substantial difference, averaging approximately 1,303.34 for Pre-1930s properties and 830.955 for Post-1930s properties. This is one of the strongest contrasts in the table and may provide useful predictive information. Lighting costs are much closer, with averages of 91.844 and 90.082, while hot-water costs are also similar, at 143.522 and 142.224. These smaller differences suggest that not every cost variable contributes equally to distinguishing the two age categories.

Overall, the summary statistics indicate that construction period is associated with meaningful differences in energy efficiency, environmental impact, energy consumption, emissions, and heating costs. Post-1930s properties generally perform better across the main energy-related measures, while Pre-1930s properties tend to consume more energy and incur higher heating costs. These patterns provide a reasonable basis for applying supervised classification models to predict whether a property belongs to the Pre-1930s or Post-1930s group.


It is always useful to view a scatterplot of the data, marking the two known groups

![](../../data_cache/vignettes/supervised_classification/scatterplot.html){width="100%" height="600px"}

The scatter plot examines the relationship between current environmental impact and current energy consumption, with observations distinguished according to whether the properties were constructed before or after 1930. The figure contains a large number of observations, allowing the overall relationship between the variables, the distributions of the two housing-age categories, and the extent of overlap between them to be examined visually.

A clear negative relationship is visible between environmental-impact score and energy consumption. As the environmental-impact score increases, current energy consumption generally decreases. Properties with lower environmental-impact scores tend to display higher and more variable energy-consumption values, whereas properties with higher environmental-impact scores are concentrated at lower consumption levels. This relationship is consistent with the interpretation that properties with stronger environmental performance typically require less energy.

The observations form a broad downward-sloping, triangular pattern. At the lower end of the environmental-impact scale, energy-consumption values are widely dispersed, ranging from comparatively low values to more than 2,000. As environmental-impact scores increase, the spread of energy consumption becomes progressively narrower. Most properties with environmental-impact scores above approximately 80 have relatively low energy-consumption values. This suggests that the variability of consumption is greater among properties with weaker environmental performance.

Differences between the two construction-period groups can also be observed. Pre-1930s properties, represented by the red points, appear throughout most of the plot and are particularly common among properties with lower environmental-impact scores and higher energy-consumption values. Post-1930s properties, represented by the blue points, appear more frequently towards the right-hand side of the plot, where environmental-impact scores are higher and energy-consumption levels are generally lower. This visual pattern supports the summary statistics, which indicated that Post-1930s properties have higher average environmental-impact scores and lower average energy consumption.

However, there is considerable overlap between the red and blue observations. Both age categories appear across much of the central region of the plot, meaning that properties from different construction periods can have similar combinations of environmental impact and energy consumption. Therefore, these two variables do not provide complete separation between the groups. This overlap is important because it suggests that classification models are unlikely to predict construction period perfectly and will require information from the additional energy-related variables.

Several extreme observations are also visible. A small number of properties have energy-consumption values considerably above the main cluster, including observations above 1,500 and approximately 2,000. Most of these extreme observations appear at low environmental-impact scores, although at least one Post-1930s observation also has exceptionally high energy consumption. These points may represent unusually inefficient properties, uncommon building characteristics, or potential data-quality issues. Nevertheless, they do not obscure the broader negative relationship.

Overall, the scatter plot demonstrates that environmental impact and energy consumption contain useful information for distinguishing between Pre-1930s and Post-1930s properties. Older properties are more frequently associated with poorer environmental performance and higher consumption, while newer properties are more concentrated around better environmental scores and lower consumption. However, the substantial overlap between the categories shows that the relationship is not sufficient for perfect classification and supports the use of multiple predictors in the subsequent supervised learning models.


## Fitting LDA and QDA classifiers

The simplest classifiers we could apply are Linear and Quadratic Discriminant analysis.


```{python}
#| echo: false
#| results: 'asis'
df = pd.read_csv('../../data_cache/vignettes/supervised_classification/lda.csv')
df.rename(columns={"Unnamed: 0": "Category"}, inplace=True)
print(df.round(3).to_markdown(index=False))
```

### Results from fitting a quadratic discriminant analysis

```{python}
#| echo: false
#| results: 'asis'
df = pd.read_csv('../../data_cache/vignettes/supervised_classification/qda.csv')
df.rename(columns={"Unnamed: 0": "Category"}, inplace=True)
print(df.round(3).to_markdown(index=False))
```

Linear Discriminant Analysis (LDA) and Quadratic Discriminant Analysis (QDA) were fitted to classify properties into two construction-period categories: Pre-30s and Post-30s. LDA produces a linear decision boundary by assuming that the two classes have a common covariance structure, whereas QDA allows each class to have a different covariance structure and can therefore produce a more flexible, quadratic decision boundary.

The test dataset contains 20,097 properties, including 10,954 Pre-30s properties and 9,143 Post-30s properties. The groups are reasonably balanced, although Pre-30s properties form a slightly larger proportion of the test observations. Consequently, accuracy remains informative, but the class-specific precision, recall and F1-scores should also be considered.

The LDA model achieved an overall classification accuracy of 0.692, meaning that approximately 69.2% of the test properties were assigned to the correct construction-period category. For Post-30s properties, the model produced a precision of 0.662, a recall of 0.662, and an F1-score of 0.662. This means that approximately 66.2% of the properties predicted as Post-30s were actually in that category, while the model also correctly identified approximately 66.2% of all true Post-30s properties.

Performance was stronger for the Pre-30s category. The LDA model achieved a precision, recall and F1-score of 0.718 for this class. Therefore, approximately 71.8% of properties predicted as Pre-30s were correctly classified, and the model identified approximately 71.8% of all true Pre-30s properties. The higher scores for the older-property class suggest that its energy-performance characteristics may be slightly easier for the model to recognise.

The QDA model produced the same reported results as LDA. Its overall accuracy was also 0.692, with precision, recall and F1-scores of 0.662 for Post-30s properties and 0.718 for Pre-30s properties. The macro-average F1-score was 0.690, while the weighted-average F1-score was 0.692 for both models. The small difference between these averages reflects the slightly unequal class sizes and the stronger performance for the larger Pre-30s group.

The identical reported metrics indicate that allowing a quadratic decision boundary did not improve predictive performance for this dataset. This may suggest that the additional flexibility of QDA was unnecessary, that the class covariance structures were sufficiently similar, or that both models produced very similar classifications. However, the reported metrics alone do not establish the exact reason.

Overall, both discriminant-analysis models demonstrate moderate predictive ability. They classify Pre-30s properties more successfully than Post-30s properties, but their accuracy of 69.2% also indicates substantial remaining overlap between the two groups. Since QDA offers no observed improvement over LDA, LDA may be preferred on grounds of simplicity, although comparison with Logistic Regression and ROC performance is required before selecting the strongest final model.


## ROC curve (and AUC) for two simple classifiers

We can of course examine the AUC for these classifiers.


### LDA ROC curve

![](../../data_cache/vignettes/supervised_classification/roc_lda.html){width="100%" height="600px"}

### QDA ROC curve

![](../../data_cache/vignettes/supervised_classification/roc_qda.html){width="100%" height="600px"}

The Receiver Operating Characteristic (ROC) curves provide a graphical assessment of the ability of the Linear Discriminant Analysis (LDA) and Quadratic Discriminant Analysis (QDA) models to distinguish between Pre-30s and Post-30s properties across a range of classification thresholds. The ROC curve plots the True Positive Rate (Sensitivity) against the False Positive Rate (1 − Specificity). A model with stronger discriminatory ability produces a curve that rises towards the upper-left corner of the graph, while the diagonal reference line represents the performance expected from random classification.

The ROC curves for both the LDA and QDA models lie above the diagonal reference line, indicating that both classifiers perform better than random guessing when distinguishing between the two construction-period categories. The curves show a similar overall shape, suggesting that the two discriminant-analysis methods produce comparable classification performance across different decision thresholds.

The Area Under the Curve (AUC) for both LDA and QDA is approximately 0.69. An AUC close to 0.70 indicates moderate discriminatory ability, meaning that the models are capable of separating Pre-30s and Post-30s properties better than chance but are not sufficiently accurate to provide near-perfect classification. This result is consistent with the classification metrics reported previously, where both models achieved an overall accuracy of 69.2%.

The similarity of the ROC curves and identical AUC values indicate that allowing a quadratic decision boundary through QDA does not improve predictive performance compared with the simpler LDA model. Despite QDA being more flexible by allowing different covariance structures for each class, both approaches produce essentially the same classification behaviour on this dataset. This suggests that the relationship between the predictor variables and housing age can be adequately represented using the assumptions of LDA.

The moderate AUC values also reflect the overlap observed in the scatter plot between the two construction-period groups. Although energy-related variables provide useful information for distinguishing older and newer properties, many observations share similar characteristics regardless of construction period. Consequently, some degree of misclassification is expected even when using supervised learning methods.

Overall, the ROC analysis confirms that both LDA and QDA provide moderate classification performance, with neither model demonstrating a clear advantage over the other. Since both methods achieve similar accuracy and AUC values, the simpler LDA model may be preferred. However, comparison with the Logistic Regression model indicates that Logistic Regression achieves higher overall classification accuracy and therefore provides the strongest predictive performance among the models considered in this coursework.

##  Additional credit


## Logistic Regression Classification Report

Logistic Regression was added as an additional supervised classification method. It models the probability of each housing-age class directly and provides a useful comparison with the existing classifiers.

```{r}
#| echo: false
#| results: asis

logistic_results <- read.csv(
  "../../data_cache/vignettes/supervised_classification/logistic.csv",
  row.names = 1
)

knitr::kable(
  logistic_results,
  digits = 3,
  caption = "Logistic Regression Classification Results"
)
```

As an additional supervised learning method, Logistic Regression was implemented to compare its performance with the Linear Discriminant Analysis (LDA) and Quadratic Discriminant Analysis (QDA) models. Unlike discriminant analysis, Logistic Regression directly models the probability that an observation belongs to a particular class and therefore does not require the stronger distributional assumptions made by LDA and QDA. This makes it a useful benchmark for evaluating whether a simpler probabilistic classification approach can improve predictive performance.

The Logistic Regression model achieved an overall classification accuracy of 74.6%, which is higher than the 69.2% accuracy obtained by both the LDA and QDA models. The weighted F1-score was 0.745, while the macro-average F1-score was 0.743, indicating balanced classification performance across both housing-age categories. For Post-30s properties, the model achieved a precision of 0.730, a recall of 0.699, and an F1-score of 0.715. For Pre-30s properties, performance was stronger, with a precision of 0.758, a recall of 0.785, and an F1-score of 0.771.

These results suggest that Logistic Regression provides a better overall fit for this dataset than the two discriminant-analysis methods. Although the model still performs slightly better when identifying Pre-30s properties, the difference between the two classes is relatively small, indicating that classification performance remains reasonably balanced. The improvement in accuracy and F1-score also suggests that directly modelling class probabilities is more effective for these energy-performance variables than relying on the assumptions required by LDA and QDA.

Overall, the additional Logistic Regression analysis demonstrates that alternative supervised learning methods can improve predictive performance while remaining computationally efficient and straightforward to interpret. Among the models considered in this coursework, Logistic Regression produced the strongest overall classification results and therefore represents the preferred model for this dataset.

## Record of AI use for MTHM503 supervised coursework

Instructions: You can use this document to record when, how and why you used GenAI to complete your assessment. It will help you create a record of AI use to submit alongside your references for AI-integrated and AI-assisted assignments. It may also be useful to help you discuss your AI use if you are required to do so in an academic conduct meeting.

| **Date**       | **AI tool used** | **Purpose**                                | **Prompt**                                 | **Hyperlink to output (where possible)** | **Section of work used for** |
|----------------|------------------|--------------------------------------------|--------------------------------------------|------------------------------------------|-------------------------------|
|22/07/2026|ChatGPT (OpenAI GPT-5)|To improve the clarity and academic style of the report and assist with interpreting model outputs. AI was also used to help identify a coding issue affecting the ROC figures.|Requested assistance with refining the introduction, interpretation of summary statistics, scatter plot, classifier results, ROC discussion, and the additional Logistic Regression section.|N/A|Introduction, Summary Statistics, Scatter Plot, LDA/QDA Results, ROC Analysis, Additional Credit|
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |
|                |                  |                                            |                                            |                                          |                               |

